---
title: 'R with Big Data 3a: Big Data and Spark SQL 1/2'
author: "Garrett Grolemund, Nathan Stephens, John Mount and Nina Zumel"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include = FALSE}
library(sparklyr)
library(dplyr)
library(nycflights13)
library(ggplot2)
```


## Flights Data

This guide will demonstrate some of the basic data manipulation verbs of dplyr by using data from the `nycflights13` R package. This package contains data for all 336,776 flights departing New York City in 2013. It also includes useful metadata on airlines, airports, weather, and planes. The data comes from the US [Bureau of Transportation Statistics](http://www.transtats.bts.gov/DatabaseInfo.asp?DB_ID=120&Link=0), and is documented in `?nycflights13`

### Connect and load data

Run the following block to connect to the cluster and copy the flights data using the `copy_to` function.

**Caveat:** The flight data in `nycflights13` is convenient for dplyr demonstrations because it is small, but in practice large data should rarely be copied directly from R objects. 

```{r message=FALSE, warning=FALSE}

sc <- spark_connect(master = "local", version = "2.0.0", hadoop_version="2.7")
flights_tbl <- spark_read_parquet(sc, "flights", "datainputs/nycflights13-parquet/flights")
airlines_tbl <- copy_to(sc, airlines, "airlines", overwrite = TRUE)

src_tbls(sc)
```

### dplyr Verbs

Verbs are dplyr commands for manipulating data. When connected to a Spark DataFrame, dplyr translates the commands into Spark SQL statements. Remote data sources use exactly the same five verbs as local data sources. Here are the five verbs with their corresponding SQL commands:

* `select` ~ `SELECT`
* `filter` ~ `WHERE`
* `arrange` ~ `ORDER`
* `summarise` ~ `aggregators: sum, min, sd, etc.`
* `mutate` ~ `operators: +, *, log, etc.`

Run the following block to see what all these commands do in Spark.

```{r}
select(flights_tbl, year:day, arr_delay, dep_delay)

filter(flights_tbl, dep_delay > 1000)

arrange(flights_tbl, desc(dep_delay))

summarise(flights_tbl, mean_dep_delay = mean(dep_delay))

mutate(flights_tbl, speed = distance / air_time * 60)
```

### Preprocess

Use filter, select, arrange, and mutate to subset the flights data. 

Fill in the ??? in the pipe below to get the departure delay, air time (in hours) and distance of all flights from the carriers United (UA), Southwest (WN), American (AA) and Delta (DL) on May 17. Arrange the results by carrier. Note that the air time as reported in column `air_time` is in minutes.

```{r}
c4  <- flights_tbl %>%
  filter(month == 5, day == 17, carrier %in% c('UA', 'WN', 'AA', 'DL')) %>%
  select(carrier, dep_delay, air_time, distance) %>%
  arrange(carrier) %>%
  mutate(air_time_hours = air_time / 60) %>%
  compute()  # this last command forces the computation and saves the result in the remote store

```

### Grouping

Used alone, the `group_by` function corresponds to the `GROUP BY` statement in SQL.

Fill in the ??? to count the number of flights and the mean departure delay by carrier.
Hint: the dplyr aggregation function `n()` counts rows and respects groups.

```{r}
c4 %>%
  group_by(carrier) %>%
  summarize(count = n(), mean_dep_delay = mean(dep_delay)) 
```

### Collecting to R : Example use

You can copy data from Spark into R's memory by using `collect()`. `collect()` executes the Spark query and returns the results to R for further analysis and visualization.

Run this block to see an example. Notice `carrierhours` is a local tbl.

```{r}
carrierhours <- collect(c4)

# Test the significance of pairwise differences and plot the results
with(carrierhours, pairwise.t.test(air_time, carrier))
ggplot(carrierhours, aes(carrier, air_time_hours)) + geom_boxplot()
```


### Spark Window Functions

dplyr implements grouped operations using Spark SQL window functions. You can compare the dplyr syntax to the query it has generated by using `show_query()` (note: this is very related to `sql_render()`).

Both the below commands use window functions, indicated by the word `PARTITION` in the SQL.
Find out which dplyr step needs window functions, by running of the pipe and looking at the SQL.

```{r, collapse=TRUE}
# Find the most and least delayed flight each day
bestworst <- flights_tbl %>%
  group_by(year, month, day) %>%
  select(dep_delay) %>% 
  filter(dep_delay == min(dep_delay) || dep_delay == max(dep_delay))
show_query(bestworst)
bestworst
```

```{r}
# Rank each flight within a daily
ranked <- flights_tbl %>%
  group_by(year, month, day) %>%
  select(dep_delay) %>% 
  mutate(rank = rank(desc(dep_delay)))
show_query(ranked)
ranked
```

### Sampling

You can use `sdf_sample` to take a random sample of rows.

Run the block below and compare the size of the sample to the size of
the original flights_tbl (Please use summarize to find the row count, as not all
remote data services implement nrow(). Spark does, but you can't always count on it).

```{r}
fsample <- sdf_sample(flights_tbl, 0.01)
```

### Writing Data

It is often useful to save the results of your analysis or the tables that you have generated on your Spark cluster into persistent storage. The best option in many scenarios is to write the table out to a [Parquet](https://parquet.apache.org/) file using the [spark_write_parquet](reference/sparklyr/spark_write_parquet.html) function. For example:

* `spark_write_parquet` Writes a Parquet file
* `spark_write_csv` Writes a CSV file
* `spark_write_json` Writes a JSON file

```{r}
spark_write_parquet(flights_tbl, "data/nycflights13-parquet/flights", mode = "overwrite")
spark_write_parquet(airlines_tbl, "data/nycflights13-parquet/airlines", mode = "overwrite")
```

### Reading Data

You can read data into Spark DataFrames using the following functions: 

* `spark_read_csv` Reads a CSV file and provides a data source compatible with dplyr
* `spark_read_json` Reads a JSON file and provides a data source compatible with dplyr
* `spark_read_parquet` Reads a parquet file and provides a data source compatible with dplyr

Regardless of the format of your data, Spark supports reading data from a variety of different data sources. These include data stored on HDFS (`hdfs://` protocol), Amazon S3 (`s3n://` protocol), or local files available to the Spark worker nodes (`file://` protocol) 

Each of these functions returns a reference to a Spark DataFrame which can be used as a dplyr table (`tbl`).

This will write the Spark DataFrame referenced by the tbl R variable to the given HDFS path. You can use the [spark_read_parquet](reference/sparklyr/spark_read_parquet.html) function to read the same table back into a subsequent Spark session:

```{r}
spark_read_parquet(sc, "data", "data/nycflights13-parquet/flights")
```

