---
title: 'R with Big Data 3b: Big Data and Spark SQL 2/2'
author: "Garrett Grolemund, Nathan Stephens, John Mount and Nina Zumel"
always_allow_html: yes
output: html_document
---

Use dplyr syntax to write Apache Spark SQL queries. Use select, where, group by, joins, and window functions in Aparche Spark SQL.

## Setup

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(sparklyr)
library(dplyr)
library(tidyr)
library(babynames)
library(ggplot2)
library(dygraphs)
library(rbokeh)
```

## Connect to Spark

Install and connect to a local Spark instance. Copy data into Spark DataFrames.

```{r}
sc <- spark_connect(master = "local", version = "2.0.0")
babynames_tbl <- copy_to(sc, babynames, "babynames", overwrite = TRUE)
applicants_tbl <- copy_to(sc, applicants, "applicants", overwrite = TRUE)
```

## Total US births

Plot total US births recorded from the Social Security Administration.

First, for practice, do it locally in the applicants data frame, and tidyr::spread

```{r, error=TRUE}
# local approach
birthsYearly <- applicants %>%  
  mutate(n_all = n_all/1000000) %>%
  tidyr::spread(sex, n_all)

# try it with remote data. Oops! This fails at tidyr::spread
birthsYearly <- applicants_tbl %>%  
  mutate(n_all = n_all/1000000) %>%
  tidyr::spread(sex, n_all)

```

Now simulate `tidyr::spread` on the remote data, and continue.

```{r}
# brutal, by-hand spread on remote data
birthsYearly <- applicants_tbl %>%
  mutate(male = ifelse(sex == "M", n_all, 0), female = ifelse(sex == "F", n_all, 0)) %>%
  group_by(year) %>%
  summarize(Male = sum(male) / 1000000, Female = sum(female) / 1000000) %>%
  arrange(year) %>%
  collect

# plot it
birthsYearly %>%
  dygraph(main = "Total US Births (SSN)", ylab = "Millions") %>%
  dySeries("Female") %>%
  dySeries("Male") %>%
  dyOptions(stackedGraph = TRUE) %>%
  dyRangeSelector(height = 20) 
```

## Aggregate data by name

Use Spark SQL to create a look up table. 

```{r}
topNames_tbl <- babynames_tbl %>%
  filter(year >= 1986) %>%  
  group_by(name, sex) %>%
  summarize(count = sum(n)) %>%
  filter(count > 1000) %>%
  select(name, sex) %>%
  compute

yearlyNames_tbl <- babynames_tbl %>%
  filter(year >= 1986) %>%
  inner_join(topNames_tbl, by = c("sex", "name")) %>%
  compute
```

## Most popular names (1986)

Identify the top 5 male and female names from 1986. Visualize the popularity trend over time.

```{r}
topNames1986_tbl <- yearlyNames_tbl %>%
  filter(year == 1986) %>%
  group_by(sex) %>%
  mutate(rank = min_rank(desc(n))) %>%
  filter(rank <= 5) %>%
  arrange(sex, rank) %>%
  select(name, sex, rank) %>%
  compute


topNames1986Yearly <- yearlyNames_tbl %>%
  inner_join(topNames1986_tbl, by = c("sex", "name")) %>%
  collect

ggplot(topNames1986Yearly, aes(year, n, color=name)) +
  facet_grid(~sex) +
  geom_line() +
  ggtitle("Most Popular Names of 1986")
```



## Shared names

Visualize the most popular names that are shared by both males and females.

```{r}
sharedName <- babynames_tbl %>%
  mutate(male = ifelse(sex == "M", n, 0), female = ifelse(sex == "F", n, 0)) %>%
  group_by(name) %>%
  summarize(Male = as.numeric(sum(male)), 
            Female = as.numeric(sum(female)),
            count = sum(n),
            AvgYear = round(as.numeric(sum(year * n) / sum(n)),0)) %>%
  filter(Male > 30000 & Female > 30000) %>%
  collect

figure(width = NULL, height = NULL, 
       xlab = "Log10 Number of Males", 
       ylab = "Log10 Number of Females",
       title = "Top shared names (1880 - 2014)") %>%
  ly_points(log10(Male), log10(Female), data = sharedName,
            color = AvgYear, size = scale(sqrt(count)),
            hover = list(name, Male, Female, AvgYear), legend = FALSE)
```
