---
title: "lazyeval"
author: "Win-Vector LLC"
date: "2/24/2017"
output: html_document
---


```{r setup}
library("sparklyr")
library("dplyr")
sc <- spark_connect(master = "local", 
                    version = "2.0.0",
                    hadoop_version="2.7")
iris_tbl <- copy_to(sc, iris, "iris", 
                    overwrite= TRUE)
```

Deliberate error (wrote "zpecies" instead of "Species", pretty much the story of my life). Yet, as we see below the error doesn't happen yet.

```{r errorButNoError}
res <- iris_tbl %>%
  group_by(zpecies) %>%
  summarize_all(funs(typical=mean))
```

But once we try to use the value we trigger the (faulty) calculation and get our error.

```{r errorFinallyShowsup, error=TRUE}
print(res)
```

One way to work around this is to force <code>compute()</code> on values you know you are going to use.

```{r compute1, error=TRUE}
res <- iris_tbl %>%
  group_by(zpecies) %>%
  summarize_all(funs(typical=mean)) %>%
  compute()
```

What a complex calculation really is:

```{r calc}
res <- iris_tbl %>%
  group_by(Species) %>%
  summarize_all(funs(typical=mean)) %>% 
  show_query
```

What a materialized table looks like:

```{r materialize}
res <- iris_tbl %>%
  group_by(Species) %>%
  summarize_all(funs(typical=mean)) %>%
  compute(name= 'summarizedIris') %>%
  show_query
```


